---
title: "Project Step 1"
author: "Minu Pabbathi"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

library(skimr)
library(tidyverse)
library(wesanderson)

data <- read.csv("~/Documents/GitHub/s24-pstat126-project/dataset/heart_disease_data_1.csv")
```

# Variable Descriptions

The `age` variable is a numeric variable that lists the age of the patient in years. The `sex` variable is a binary variable that states the sex of the patient, with 0 representing female and 1 representing male. The `chest.pain.type` variable is a categorical variable that classifies the type of chest pain the patient is experiencing; 0 represents typical angina, 1 represents atypical angina, 2 represents non-anginal pain, and 3 represents asymptomatic. The `resting.bp.s` variable is a numeric variable that states the resting blood pressure in mmHg.

# Summary Statistics

## Age

```{r}
summary(data$age)
```

The ages of the patients range from 29 to 77. The average age is 54.02 and the median age is 55.

## Resting Blood Pressure

```{r}
summary(data$trestbps)
```

The resting blood pressure ranges from 94 to 200. The average resting blood pressure is 132.5 mmHg and the median is 130 mmHg.

# Graphical Representations

## Age

```{r}
hist(data$age, xlab = "Age", main = "Age of Patients", col = "thistle")
```

Based on the histogram, the most common age group appears to be 55-60. The graph is slightly skewed to the left and the data is unimodal.

## Sex

```{r}
barplot(table(data$sex), xlab = "Sex", main = "Sex of Patients", col = wes_palette("GrandBudapest2"))
```

This barplot indicates that there are more men in the study than women.

## Chest Pain Type

```{r}
barplot(table(data$cp), xlab = "Chest Pain Type", main = "Number of Reports of Each Chest Pain Type", names.arg = c("typical angina", "atypical angina", "non-anginal pain", "asymptomatic"), col = wes_palette("Royal2", 4))
```

The barplot indicates that majority of patients experienced typical angina, followed by non-anginal pain, atypical angina, and asymptomatic patients.

## Resting Blood Pressure

```{r}
boxplot(data$trestbps, horizontal = T, main = "Resting Blood Pressure", xlab = "Blood Pressure in mmHg", outline = F, col = "lightblue")
```

The boxplot (outliers excluded) indicates that median resting blood pressure is around 130 and the data is evenly spread.
