---
title: "Project Step 1"
author: "Minu Pabbathi"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

library(skimr)
library(tidyverse)
library(wesanderson)

data <- read.csv("~/Documents/GitHub/s24-pstat126-project/dataset/heart_disease_data.csv")
```

# Variable Descriptions

The `age` variable is a numeric variable that lists the age of the patient in years. The `sex` variable is a binary variable that states the sex of the patient, with 0 representing female and 1 representing male. The `chest.pain.type` variable is a categorical variable that classifies the type of chest pain the patient is experiencing; 1 represents typical angina, 2 represents atypical angina, 3 represents non-anginal pain, and 4 represents asymptomatic. The `resting.bp.s` variable is a numeric variable that states the resting blood pressure in mmHg.


# Summary Statistics

## Age

```{r}
summary(data$age)
```

The ages of the patients range from 28 to 77. The average age is 51.55 and the median age is 52.

## Resting Blood Pressure

```{r}
summary(data$resting.bp.s)
```

The resting blood pressure ranges from 0 to 200. The average resting blood pressure is 132 mmHg and the median is 130 mmHg.

# Graphical Representations

## Age

```{r}
hist(data$age, xlab = "Age", main = "Age of Patients", col = "thistle")
```

## Sex

```{r}
barplot(table(data$sex), xlab = "Sex", main = "Sex of Patients", col = wes_palette("GrandBudapest2"))
```

## Chest Pain Type

```{r}
barplot(table(data$chest.pain.type), xlab = "Chest Pain Type", main = "Number of Reports of Each Chest Pain Type", names.arg = c("typical angina", "atypical angina", "non-anginal pain", "asymptomatic"), col = wes_palette("Royal2", 4))
```


## Resting Blood Pressure

```{r}
boxplot(data$resting.bp.s, horizontal = T, main = "Resting Blood Pressure", xlab = "Blood Pressure in mmHg", outline = F, col = "lightblue")
```

