---
title: "Project Step 1"
author: "Minu Pabbathi, Wendy Zhao, Paul Zhang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

library(skimr)
library(tidyverse)
```

# Variable Descriptions

The variable `school` is a binary variable that states which school the student attends ('GP' or 'MS').

The `sex` variable is a binary variable that lists the sex of the student ('M' or 'F').

The `age` variable is numerical and lists the student's age from 15-22.

The `address` variable is categorical and states whether the student lives in a rural (R) or urban (U) area.

The `famsize` variable is a categorical variable that states whether the student's family is less than or equal 3 (LE3) or greater than 3 (GT3).

`Pstatus`is a categorical variable that classifies parental cohabitation as living together (T) or living apart (A).

`Medu` is a categorical variable that lists the student's mother's education level on a scale from 0-4. Similarly, the `Pedu` is a categorical variable that lists the student's father's education level on a scale from 0-4. 0 indicates none, 1 indicates primary education (4th grade), 2 indicates 5th to 9th grade, 3 indicates secondary education, and 4 indicates higher education.

`Mjob` is a categorical variable that lists the student's mother's job. `Fjob` is similar but refers to the student's father's job. The categories include 'teacher', 'health' care related, civil 'services' (e.g. administrative or police), 'at_home', or 'other'.

The variable `reason` is a categorical variable that lists the reason why the student decided to attend the school. The categories include close to 'home', school 'reputation', 'course' preference, or 'other'.

# Data Summary

```{r}
student_data <- read.csv("~/Documents/GitHub/s24-pstat126-project/student/student-mat.csv", header = T)
skim(student_data)

```

# Graphs

## Sex

```{r}
barplot(table(student_data$sex), xlab = "Sex", main = "Number of Students by Sex")
```


## Address

```{r}
barplot(table(student_data$address), xlab = "Rural or Urban Area", ylab = "Frequency", main = "Students Living in Rural Areas vs. Urban Area")
```

## Parents' Jobs

```{r}
mjob <- table(student_data$Mjob)
fjob <- table(student_data$Fjob)

jobs <- rbind(mjob, fjob)

barplot(jobs, beside = T, legend.text = c("Mother's Job", "Father's Job"), xlab = "Type of Job", main = "Mother's vs. Father's Job")
```

## Reason For Attending School

```{r}
barplot(table(student_data$reason), main = "Reasons for Choosing School", xlab = "Reason")
```

## Parents' Education

```{r}
medu <- table(student_data$Medu)
fedu <- table(student_data$Fedu)

edu <- rbind(medu, fedu)

barplot(edu, beside = T, legend.text = c("Mother's Education", "Father's Education"), xlab = "Level of Education", main = "Mother's vs. Father's Education")
```

