---
title: "Project Step 1"
author: "Minu Pabbathi, Ziqian Zhao, Paul Zhang"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = F,
                      results = 'markup',
                      fig.width = 5,
                      fig.height = 3,
                      fig.align = 'center',
                      message = F,
                      warning = F)

# packages
library(tidyverse)
library(faraway)
library(skimr)
library(readr)

heart_disease_data_1 <- read.csv("~/Documents/GitHub/s24-pstat126-project/dataset/heart_disease_data_1.csv")
```


# About the data

This data set dates from 1988 and consists of four databases: Cleveland, Hungary, Switzerland, and Long Beach V. It contains information about patients and risks for heart disease as well as whether or not they eventually end up developing heart disease. The data set originally contains 1025 observations. For the sake of clearness of the scatter plot, we randomly choose 500 from the original dataset.


# Data Description

The `age` variable is a numeric variable that lists the age of the patient in years. The `sex` variable is a binary variable that states the sex of the patient, with 0 representing female and 1 representing male. The `chest.pain.type` variable is a categorical variable that classifies the type of chest pain the patient is experiencing; 0 represents typical angina, 1 represents atypical angina, 2 represents non-anginal pain, and 3 represents asymptomatic. The `resting.bp.s` variable is a numeric variable that states the resting blood pressure in mmHg.

The `cholesterol` variable is a numeric data which give an overview of person's cholesterol levels. The `fasting blood sugar` variable is a binary data with 1 means the fast blood sugar is greater than 120mg/dl, and 0 means the fast blood sugar is not. The `resting electrocardigram` variable is a nominal data with scale of 0 (normal), 1 (abnormality in ST-T waves), and 2 (show left ventricular hypertrophy), which record the resting electrocardiogram results. The `maximum heart rate` variable is a numeric data which records the maximum heart rate achieved.

The variable `exang` indicates that if the patient have exercise induced angina (chest pain), which is shown with binary data "yes" or "no". The variable `oldpeak` is a float type data records the ST depression, which is a measure of abnormality of an electrocardiogram, and the measurement is in unit depression. The variable `ST slope` is the slope of the peak exercise ST segment, which is an electrocardiography read out indicating quality of blood flow to the heart, and the data are in nominal type including 0 (upsloping), 1 (flat) and 2 (downsloping). Finally, the variable `target` is a binary data showing whether the patient is having heart disease (1 = having heart disease and 0 = normal).

# Summary Statistics and Graphs

## Age

```{r}
summary(heart_disease_data_1$age)
```

The ages of the patients range from 29 to 77. The average age is 54.02 and the median age is 55.

## Resting Blood Pressure

```{r}
summary(heart_disease_data_1$trestbps)
```

The resting blood pressure ranges from 94 to 200. The average resting blood pressure is 132.5 mmHg and the median is 130 mmHg.

## Cholesterol

```{r}
summary(heart_disease_data_1$chol)
```

## Resting Electrocardiogram

| Nominal | 0 | 1 | 2|
|:-------|:-------:|:------:|------:|
| Meaning |   normal   |  abnomality in ST-T waves   | Show left ventricular hypertrophy| 
| No. of people |263|233|4|

## Max Heart Rate

This data is a numeric dataset which records the maximum heart rate achieved.

```{r}
summary(heart_disease_data_1$thalach)
```

## Chest Pain Type

```{r}
barplot(table(heart_disease_data_1$cp), xlab = "Chest Pain Type", main = "Number of Reports of Each Chest Pain Type")
```


| 0 | 1 | 2 | 3 |
----|---|----|-------
|typical angina | atypical angina | non-anginal pain | asymptomatic|

The barplot indicates that majority of patients experienced typical angina, followed by non-anginal pain, atypical angina, and asymptomatic patients.

## Resting Blood Pressure

```{r}
boxplot(heart_disease_data_1$trestbps, horizontal = T, main = "Resting Blood Pressure", xlab = "Blood Pressure in mmHg", outline = F)
```

The boxplot (outliers excluded) indicates that median resting blood pressure is around 130 and the data is evenly spread.

## Fasting Blood Sugar

```{r}
barplot(table(heart_disease_data_1$fbs), 
        xlab = "Fasting blood sugar", main = "Fasting blood sugar",
        names.arg = c("<= 120 mg/dl", "> 120 mg/dl"))
```

## Exercise induced angina

```{r}
barplot(table(heart_disease_data_1$exang), ylab = "Number of patients", main="Exercise induced angina",
        names.arg = c("No", "Yes"))
```

## Target

```{r}
barplot(table(heart_disease_data_1$target), ylab = "Number of patients", main="Heart disease", names.arg = c("Normal", "Has Heart Disease"))
```

