---
title: "PSTAT 126 Project Step 1"
author: "Ziqian Zhao"
date: '2024-04-17'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = F,
                      results = 'markup',
                      fig.width = 4,
                      fig.height = 3,
                      fig.align = 'center',
                      message = F,
                      warning = F)

# packages
library(tidyverse)
library(faraway)
```

```{r}
load("student_mat.RData")
```
## Data Description

Guardian (Categorical Variable)

This data describe the guardian of the students in both school.

| Category | mother | father | other |
|:---------|:------:|:------:|:-----:|
| Number   |  273   |   90   |  32   |

Travel Time (Integer Variable)

This data describe students' travel time from home to school.

| Scale (t in minutes)  | 1 (t \< 15 ) | 2 ( 15 \< t \< 30) | 3 ( 30 \< t \< 60 ) | 4( t \> 60) |
|:--------------|:-------------:|:-------------:|:-------------:|:-------------:|
| Number |       257       |            107             |             23             |        8        |

Summary:

```{r}
summary(student_mat$traveltime)
```

Study Time (Integer Variable) 

This data describes students' weekly study time.


| Scale (t in hours) | 1 (t \< 2) | 2 (2 \< t \< 5 ) | 3 (5 \< t \< 10 ) | 4 (t \> 10 ) |
|:--------------|:-------------:|:-------------:|:-------------:|:-------------:|
| No. of students |      105       |         198         |          65          |       27        |

```{r}
summary(student_mat$studytime)
```

Failures(Integer Variables)

This data describes number of non-past classes

| Scale  |  0  |  1  |  2  |  3  | more than 3 |
|:-------|:---:|:---:|:---:|:---:|:-----------:|
| No. of  | 312 | 50  | 17  | 16  |      0      |

```{r}
summary(student_mat$failures)
```

SchoolSup (Binary Variable)

This data desbribe whether students have extra educational support

| Binary | 1 (yes) | 0 (no) |
|:-------|:-------:|:------:|
| Number |   51    |  344   |

FamSup (Binary Variable)

This data describe whether student have family educational support.

| Binary | 1 (yes) | 0 (no) |
|:-------|:-------:|:------:|
| Number |   242   |  153   |

Paid (Binary Variable)

This data describe whether student taking extra paid class for math study.

| Binary | 1 (yes) | 0 (no) |
|:-------|:-------:|:------:|
| Number |   181   |  214   |

Activities (Binary Variable)

This data describes whether students have extra-curricular activities.

| Binary | 1 (yes) | 0 (no) |
|:-------|:-------:|:------:|
| Number |   201   |  194   |

Nursery (Binary Variable)

This data describes whether students had attended a nursery school

| Binary | 1 (yes) | 0 (no) |
|:-------|:-------:|:------:|
| Number |   314   |   81   |

Higher (Binary Variable)

This data describe whether students want to take higher education

| Binary | 1 (yes) | 0 (no) |
|:-------|:-------:|:------:|
| Number |   375   |   20   |

Internet (Binary Variable)

This Data describe whether student can access internet at home.

| Binary | 1 (yes) | 0 (no) |
|:-------|:-------:|:------:|
| Number |   329   |   66   |

## Binary Relationship

