---
title: "pstats126 -Step1"
author: "Ziqian Zhao"
date: '2024-04-23'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = F,
                      results = 'markup',
                      fig.width = 5,
                      fig.height = 3,
                      fig.align = 'center',
                      message = F,
                      warning = F)

# packages
library(tidyverse)
library(faraway)
library(skimr)
library(readr)
```

```{r}
heart_disease_data_1 <- read_delim("s24-pstat126-project/dataset/heart_disease_data_1.csv", 
                                   delim = ",", escape_double = FALSE, trim_ws = TRUE)
```
# About the data

This data set dates from 1988 and consists of four databases: Cleveland, Hungary, Switzerland, and Long Beach V. It originally contains 76 attributes, including the predicted attribute, but all published experiments refer to using a subset of 14 of them. The data set originally contains 1025 observations. For the sake of clearness of the scatter plot, we randomly choose 500 from the original dataset. 

# Data Description

## Cholesterol 

This data is a numeric data which give an overview of person's cholesterol levels.
```{r}
summary(heart_disease_data_1$chol)
```
```{r}
barplot(table(heart_disease_data_1$chol), 
        xlab = "Cholesterol", main = "Cholesterol Level")
```

## Fasting blood sugar

This data is a binary data with 1 means the fast blood sugar is greater than 120mg/dl, and 0 means the fast blood sugar is not.

| Binary | 1 (> 120 mg/dl) | 0 ($\leq$ 120mg/dl) |
|:-------|:-------:|:------:|
| Number |   82   |  418   |

```{r}
barplot(table(heart_disease_data_1$fbs), 
        xlab = "Fasting blood sugar", main = "Fasting blood sugar")
```

## Resting Electrocardiogram 

This data is a nomial data with scale of 0, 1, 2, which record the resting electrocardiogram results

| Nominal | 0 | 1 | 2|
|:-------|:-------:|:------:|------:|
| Meaning |   normal   |  abnomality in ST-T waves   | Show left ventricular hypertrophy| 
| No. of people |263|233|4|

```{r}
barplot(table(heart_disease_data_1$restecg), xlab = "Scale of Resting ECG", main = "Resting ECG")
```
## Max Heart Rate

This data is a numeric dataset which records the maximum heart rate achieved.

```{r}
summary(heart_disease_data_1$thalach)
```

# Binary Raltionship

The graph examines the relationship between serum cholestoral and maximum hearting rate. From the graph, we cannot conclude a linear relationship between these two variables. 

```{r}
plot(x=heart_disease_data_1$chol, y=heart_disease_data_1$thalach,
     xlab = "Serum cholestoral in mg/dl",
     ylab = "Maximum hearting rate achieved",
     main = "Relationship between chol and heart rate")
```

