---
title: "Project Step2"
author: "Ziqian Zhao"
date: "2024-05-03"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# knit options
knitr::opts_chunk$set(echo = F,
                      results = 'markup',
                      message = F,
                      warning = F)


# packages
library(tidyverse)
library(faraway)
library(skimr)
library(readr)
library(plotly)
library(readxl)
heart_disease_data_1 <- read.csv("dataset/heart_disease_data_1.csv")
```

# Introuction

# Scatter Plot

```{r,fig.width=6, fig.height=6, fig.align='center', fig.cap='Fig.1 Relation between age and maximum heart rate achieve'}

p <- ggplot(heart_disease_data_1, aes(x = age, y = thalach)) + geom_point() +
  labs(x = 'Age', y = 'Max Heart Rate') + 
  geom_smooth(method = 'lm', 
        formula = 'y ~ x',
        se = F)
p
```
# Hypothesis Test

$$
H_0: \beta_1 = 0
$$
$$
H_1: \beta_1 \neq 0
$$

# Analysis of p-value and confidence interval

## p-value
```{r}
fit <- lm(formula = thalach ~ age, data = heart_disease_data_1)
summary(fit)
```



From the summary table, we summarized the model as 
$$
\text{Heart Rate}_i = 206 - 1.03\cdot\text{Age}_i
$$
The p-value for $\beta_1 < 2\times10^{-16}$, so we reject null hypothesis and conclude that there is association between age and maximum heart rate achieved.


## Construct Confidence interval

```{r}
b_bar = fit$coefficients[[2]]
t = qt(p=0.025, df = 498, lower.tail = FALSE)
conf.interval <- c(b_bar-0.1034*t, b_bar+0.1034*t)
conf.interval
```
The confidence interval for $\beta_1$ is (-1.24,-0.83) at a significant level of 95%. This is to say that repeatedly estimating $\beta_1$ with same procedure, 95% of the estimation lies between this interval. Since 0 is not included in this interval, it is most likely that $\beta_1 \neq 0$. Therefore, we conclude that there is some association between age and maximum heart rate achieved.
